Statically Typed String Sanitation Inside a Python

### 2014-08-07 sanitation-psp14/sanitation-psp14-short.tex
changes

\newcommand{\val}{{\sf val}}

@@

% Meta-theoretic functions
\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
Web applications must ultimately generate strings that contain commands to be consumed by systems like web browsers and database engines. If these strings are constructed from user input that has not been properly sanitized, this can expose costly injection vulnerabilities. %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.

In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language. The language of a string is tracked by the type system through operations like concatenation, substitution and coercion, so regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that constructs  commands. Simple type annotations at key points can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks.

We take the position that to be practical, such a type system cannot require that programmers adopt a new programming language. Ideally, such a ``special-purpose'' type system would be distributed as a library that could safely be used together with other such type systems. We support this by specifying a sound translation to a simple language containing only strings and regular expressions, then discuss implementing the type system together with this translation as a library in \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Improper input sanitation is a leading cause of security vulnerabilities in web applications \cite{OWASP}. Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command. Modern web frameworks, libraries, and database abstraction layers attempt to ensure proper sanitation of user input. When these methods are unavailable or insufficient, developers implement custom sanitation techniques. In both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.

@@

To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions. If $s : \stringin{r}$, then $s$ is a string matching the language $r$. Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$. The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$. Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted). We also show how this system is translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.

@@

A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities \cite{UrFlowOSDI10}. 
Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.

@@

We are also unaware of any extensible programming languages which emphasize applications to security concerns.

Incorporating regular expressions into the type system is not novel. The XDuce system \cite{HosoyaVouillonPierce2000ICFP,HosoyaPierce2002} checks XML documents against schemas using regular expressions. 
Similarly, XHaskell \cite{xhaskell} focuses on XML documents.
We differ from this and related work in at least two ways. First, our system  is defined within an extensible type system; second, and more importantly, we have demonstrated that regular expression types are applicable to the web security domain.

@@

In this section we define a language for statically checked string sanitation ($\lambdas$).
The system has regular expression types $\stringin{r}$ whre $r$ is a regular expression. 
Expressions of this type evaluate to string literals in the language described by $r$.
Operations on expressions of type $\stringin{r}$ preserve this property; for instance,
the result of string concatenation is

The premier operation for manipulating strings in $\lambdas$ is string substitution, 
which is a familiar operation to any programmer who has used regular expressions.
The replacement operation replaces all instances of a pattern in one string with
another string; for instance, $\lsubst(a|b, ``a", ``"c") = c$.
In order to computer the type resulting from aubstitution, we also need to compute
the result of replacing on language with another inside a given language.
Finally, just for convienance, we provide a coerce operation. The introduction of
coercion requires handling of runtime errors.

The underlying language $\lambdap$ has only one type for strings. We prove that whenever
a term is translated from $\lambdas$ to $\lambdap$, correctness is preserved.
The only exception is in the case of unsafe casts in $\lambdas$, which are unnecessary but
are included to demonstrate that the regex library of $\lambdap$ may be used to insert
dynamic checks whenever even when developers are not careful about using statically checked
operations.

A brief outline of this section follows:
\begin{itemize}
  \item Page 3 contains a definition of $\lambdas, \lambdap$ and the translation from $\lambdas$ to $\lambdap$.
  \item In \S 2.1 we state some properties about regular expressions which are needed in our correctness proofs.
   \item In \S 2.2 we prove type safety for $\lambdap$ as well as both type safety and correctness for $\lambdas$.
  \item In \S 2.3 we prove that translation preserves the correctness reesult about $\lambdas$.
\end{itemize}

%The $\lambdas$ language is characterized by a type of strings indexed by regular
%expressions, together with operations on such strings which correspond to common
%input sanitation patterns.
%This section presents the grammar, typing rules and operational semantics for
%$\lambdas$ as well as an underlying language $\lambdap$.
%
%The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
%types ensuring a string belongs to a specified language. For instance, 
%$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
%the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
%Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.
%
%The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
%Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
%In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).
%
%Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$. Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.
%
%Unfortunately, we are unable to present full proofs in this paper due to space constraints; however, proofs are given in full in the accompanying technical report?.
%%Note that we never insert an internal check where the type of the string implies
%%that a check must succeed. Furthermore, expensive calculations (such as language
%%inclusion or ${\sf rreplace}$) all occur at compile time.
%%Since sanitation problems are generally expressible with smalle, simple expressions,
%%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
%%support this assertion.

@@

\begin{figure}[h]
\begin{grammar}
<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

<$\psi$> ::=	$\psi \rightarrow \psi$			\hfill	source types					\alt
$\stringin{r}$

<S> ::= 
      $\lambda x . e$ \hfill source terms \alt
      $e e$ \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{S}{S}$ \alt
      $\sisubst{r}{S}{S}$ \alt
      $\coerce{r}{S}$
\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
\end{grammar}
\end{figure}

@@

<$\theta$> ::= $\theta \rightarrow \theta$ \hfill target types \alt
$\str$ \alt $\regex$

<P> ::= $\lambda x . e$ \hfill target terms \alt
  $e e$ \alt
  $\tstr{s}$ \alt
  $\rx{r}$ \alt
  $\tconcat{P}{P}$ \alt
  $\preplace{P}{P}{P}$ \alt
  $\tcheck{P}{P}$ 
  \end{grammar}
\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

@@

% The translations rules *just won't* go to the bottom of the page, so 
% instead I am placing them at the top. Since the figures are out of order,
% it's even more important that we get the number gith. therefore, the counter
% is modified before and after this figure.

\addtocounter{figure}{4}
\begin{figure*}\label{fig:trans}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
\begin{mathpar}
%\inferrule[S-T-Abs]
%{\sctx, x : T_1 \vdash e : T_2}
%{\sctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[S-T-App]
%{\sctx \vdash e_1 : T' \\ \sctx \vdash e_2 : T' \rightarrow T}
%{\sctx \vdash e_1 e_2 : T}
%  
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-Replace]
{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ \lreplace{r}{r_1}{r_2} = r'}
{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}

@@

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
%
%\inferrule[S-E-Abs]                                                             
%{ \ }                                                                           
%{\lambda x.e \sreduces \lambda x.e}                                             
%
%\inferrule[S-E-App]
%{ e_1 \sreduces e_1' \\ e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-Replace]
{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \lsubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}

@@

\begin{mathpar}
%\inferrule[P-T-Abs]
%{\tctx, x : T_1 \vdash e : T_2}
%{\tctx \vdash \lambda x.e : T_1 \rightarrow T_2}
%
%\inferrule[P-T-App]
%{\tctx \vdash e_1 : T \\ \tctx \vdash e_2 : T' \rightarrow T}
%{\tctx \vdash e_1 e_2 : T}
%
\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\begin{mathpar}

%\inferrule[P-E-Abs]
%{ \ }
%{\lambda x.e \sreduces \lambda x.e}
%
%\inferrule[P-E-App]
%{ e_1 \sreduces e_1' \\  e_2 \sreduces e_2' }
%{ e_1 e_2 \sreduces e_1' e_2'}
%
  \inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Replace]
{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \lsubst{r}{s_2}{s_3} = s} 
{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}

@@

Our type safety proof for language S replies on a relationship between
string substitution and language substitution given in lemma \ref{thm:substcorrespondence}.
We also rely upon several other properties of regular languages.
Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
language recognized by the regular expression $r$. This distinction between the regular expression
and its language -- typically elided in the literature -- makes our definition and
proofs about systems S and P more readable.

\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
The following are properties of regular expressions which are necessary for our proofs:
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
For all strings $s$ and regular expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
Regular languages are closed under difference, right quotient, reversal, and string homomorphism.
\end{lem}

If any of these properties are unfamiliar, the reader may refer to a standard text on the subject \cite{cinderella}.

\begin{defn}[$\tt{lsubst}$]
The replation $\lsubst{r}{s_1}{s_2} = s$ produces a string $s$ in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
  The relation $\lreplace{r}{r_1}{r_2} = r'$ relates $r, r_1,$ and $r_2$ to
  a language $r'$ containing all strings of $r_1$ except that any substring $s_{pre} s s_{post} \in \lang{r_1}$ where $s \in \lang{r}$
  is replaced by the set of strings $s_{pre} s_2 s_{post}$ for all $s_2 \in \lang{r_2}$ (the prefix and postfix positions may be empty).
\end{defn}

\begin{lem}{Closure.} \label{thm:total}
  If $\lang{r}, \lang{r_1}$ and $\lang{r_2}$ are regular expressions, then $\lang{\lreplace{r}{r_1}{r_2}}$ is also a regular language.
\end{lem}
\begin{proof}
The theorem follows from closure under difference, right quotient, reversal and string homomorphism.
\end{proof}

\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
  If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{s_1}{s_2}}$.
\end{lem}
\begin{proof}
The theorem follows from the definitions of lsubst and lreplace; note that language substitutions over-approximate string substititons.
\end{proof}

%\begin{description}
%\item[case $r=\alpha$.]
%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
%
%\item[case $r = a|b$].
%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
%
%\item[case $r = ab$].
%By a similar argument to the disjunctive case.
%
%\item[case $r = a*$].
%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
%\end{description}

@@

\begin{lem}
  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression.
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
\end{proof}
\begin{lem}
If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\end{lem}

@@

\begin{thm}{Safety and Sanitation Correctness for the String Fragment of P.} \label{thm:scorrect}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
  then $S \sreduces \rstr{s}$ and $\sctx \vdash \rstr{s} : \stringin{r}$; or else $S \err$.
\end{thm}
\begin{proof}
By induction on the typing relation, where (a) case holds by lemma \ref{thm:regexprops} in the S-T-Concat case and lemma \ref{thm:substcorrespondence} in the S-T-Replace case..
The (b) cases hold by unstated, but standard, error propagation rules.
\end{proof}

In addition to safety, we proof a correctness result for $\lambdas$ which ensures that well-typed terms of regular string type are in the language associated with their type.

\begin{thm}{Correctness of Input Sanitation for $\lambdas$.}\label{thm:scorrect}
  If  $\sctx \vdash S : \stringin{r}$ and $S \sreduces \rstr{s}$ then $s \in \lang{r}$.
\end{thm}
\begin{proof}
  Follows directly from type safety, canonical forms for $\lambdas$.
\end{proof}

%\begin{thm}{Correctness of Input Sanitation (CIS) for $\lambdas$.}\label{thm:scorrect}
%  If $\sctx \vdash S : \stringin{r}$ then either (a) $S \sreduces \rstr{s}$ for $s \in \lang{r}$
%  or (b) $S \err$.
%\end{thm}

\begin{thm} 
  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
  then $P \sreduces P'$ and $\tctx \vdash P' : \theta$, or else $P \err$.
\end{thm}

>>>>

@@

% Meta-theoretic functions
\newcommand{\dosubst}[3]{{\tt replace}(#1,#2,#3)}
\newcommand{\lsubst}[3]{{\tt lsubst}(#1,#2,#3)} % This used to render lreplace(...) so there're probably mistkes wherever this command was used now.
\newcommand{\lreplace}[3]{{\tt lreplace}(#1, #2, #3)}

@@

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
Web applications must ultimately generate strings containing commands that are   executed by external systems like web browsers and database engines. If these strings are constructed from user input that has not been properly sanitized, this can lead to a variety of command injection vulnerabilities. %Structured frameworks for constructing commands centralizes control over string sanitization, but verifying that the framework itself sanitizes user input correctly can still be tricky.

In this paper, we introduce \emph{regular string types}, which classify strings known statically to be in a specified regular language and support operations like concatenation, substitution and coercion. Regular string types can be used to implement, in essentially a conventional manner, the parts of a web application or application framework that construct such command strings. Straightforward type annotations at key interfaces can be used to statically verify that sanitization has been performed correctly without introducing redundant run-time checks. We specify this type system as a minimal typed lambda calculus, $\lambda_{RS}$.

To be practical, adopting a type system like this should not require adopting a new programming language. Instead, we favor extensible type systems: new static type systems like this should be distributed as libraries atop a mechanism that guarantees that they can be safely composed.  We support this by 1) specifying a  translation from $\lambda_{RS}$ to a language containing only strings and regular expressions, then, taking Python as such a language, 2) implement the type system together with the translation as a library using \texttt{atlang}, an extensible static type system for Python (being developed by the authors).% represents the most sophisticated application of this library to date.
% Security-oriented programming languages do not enjoy wide adoption. Conversely, library and framework-based approaches to security are pervasive. Unfortunately, libraries and frameworks are only effective solutions if they are implemented and used properly. In this paper, we propose that extensible programming languages provide a promising mechanism for increasing the amount of trust that can be place in libraries and frameworks. To demonstrate this approach, we present a conservative extension to the simply-typed lambda calculus for checking the correctness of input sanitation algorithms. We also demonstrate how this language's security guarantees are preserved under translation into an underlying language containing a regular expression library. We believe this approach -- complementing existing techniques with light-weight, composable type-based analyses -- constitutes a promising compromise between usability, potential for industry adoption, and theoretically grounded safety guarantees.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Improper input sanitation is a leading cause of security vulnerabilities in web applications [OWASP]. Command injection attacks exploit improper input sanitation by inserting malicious code into an otherwise benign command. Modern web frameworks, libraries, and database abstraction layers attempt to ensure proper sanitation of user input. When these methods are unavailable or insufficient, developers implement custom sanitation techniques. Im both cases, sanitation algorithms are implemented using the language's regular expression capabilities and usually \emph{replace} potentially unsafe strings with equivalent escaped strings.

@@

To demonstrate this approach, we present a simply typed lambda calculus with \emph{constrained strings}; that is, a set of string types parameterized by regular expressions. If $s : \stringin{r}$, then $s$ is a string matching the language $r$. Additionally, we include an operation $\rreplace{r}{s_1}{s_2}$ which corresponds to the replace mechanism available in most regular expression libraries; that is, any substring of $s_1$ matching $r$ is replaced with $s_2$. The type of this expression is the computed, and is likely ``smaller" or more constrained than the type of $s_1$. Libraries, frameworks or functions which construct and execute commands containing input can specify a safe subset $\stringin{r_{\sf spec}}$ of strings, and input sanitation algorithms can construct such a string using ${\sf rreplace}$ or, optionally, by coercion (in which case a runtime check is inserted). We also show how this simple can be translated into a host language containing a regular expression library such that the safety guarantee of the extended language is preserved.

@@

A number of research languages provide static guarantees that a program is free of input sanitation vulnerabilities [Jif][Ur/Web]. 
Unlike this work, our solution to the input sanitation problem has a very low barrier to adoption; for instance, our implementation conservatively extends Python -- a popular language among web developers.
We also believe our general approach is better-positioned for security, where continuously evolving threats might require frequent addition of new analyses; in these cases, the composability and generality of our approach is a substantial advantage.

@@

We are also unaware of any extensible programming languages which emphasize applications to security concerns (TRUE?).

Incorporating regular expressions into the type system is not novel. The XDuce system \cite{pierce} typechecks XML schemas using regular expressions. We differ from this and related work in at least two ways. First, our system  is defined within an extensible type system; second, and more importantly, we have demonstrated that regular expression types are applicable to the web security domain.

@@

The $\lambdas$ language is characterized by a type of strings indexed by regular
expressions, together with operations on such strings which correspond to common
input sanitation patterns.
This section presents the grammar, typing rules and operational semantics for
$\lambdas$ as well as an underlying language $\lambdap$.

The system $\lambdas$ is the simply typed lambda calculus extended with \emph{regular expression types}, which are string
types ensuring a string belongs to a specified language. For instance, 
$S : \stringin{r}$ reads ``$s$ is a string matching $r$".
the system includes an operation for replacing all instances of a pattern $r$ in a string $s_1$ with another string $s_2$. 
Input sanitation algorithms -- as implemented by developers or within popular libraries and frameworks -- are often implemented in terms of this replace operation.

The language $\lambdap$ is a simple functional language extended with a minimal regular expression library.
Any general purpose programming language could stand in for $\lambdap$; for instance, SML or Python.
In an implementation, our correctness results are modulo the underlying language's correct implementation of regular expression matching (see P-E-Replace).

Finally, we define a translation from our type system $\lambdas$ into $\lambdap$ which preserves the safety guarnatee codified in the string types of $\lambdas$. Because our extension is conservative and its guarnatees are preserved under translation to an underlying language, it is highly composable with other analyses.

Unfortunately, we are unable to present full proofs here due to space constraints.
%Note that we never insert an internal check where the type of the string implies
%that a check must succeed. Furthermore, expensive calculations (such as language
%inclusion or ${\sf rreplace}$) all occur at compile time.
%Since sanitation problems are generally expressible with smalle, simple expressions,
%we believe that the compile-time overhead is not significant enough to prohibiour target usecase. Informal experimentation with our implementation seems to
%support this assertion.

@@

\begin{figure}[t]
\begin{grammar}
<r> ::= $\epsilon$ | $.$ | $a$ | $r \cdot r$ | $r + r$ | $r*$ \hfill $a \in \Sigma$

@@

<$\psi$> ::=	...			\hfill	source types					\alt
$\stringin{r}$

<S> ::= ... \hfill source terms \alt
      $\strin{s}$ \hfill $s \in \Sigma^{*}$ \alt
      $\rsconcat{S}{S}$ \alt
      $\sisubst{r}{S}{S}$ \alt
      $\coerce{r}{S}$
\caption{Syntax for the string sanitation fragment of our source language, $\lambdas$.}
\end{grammar}
\end{figure}

@@

<$\theta$> ::= ... \hfill target types \alt
$\str$ \alt $\regex$

<P> ::= ... \hfill target terms \alt
  $\tstr{s}$ \alt
  $\rx{r}$ \alt
  $\tconcat{P}{P}$ \alt
  $\preplace{P}{P}{P}$ \alt
  $\tcheck{P}{P}$ 
  \end{grammar}
\caption{Syntax for the fragment of our target language, $\lambdap$, containing strings and statically constructed regular expressions.}
\label{fig:lcsSyntax}
\end{figure}

@@

% The translations rules *just won't* go to the bottom of the page, so 
% instead I am placing them at the top. Since the figures are out of order,
% it's even more important that we get the number gith. therefore, the counter
% is modified before and after this figure.
\addtocounter{figure}{4}
\begin{figure*}\label{fig:trans}

@@

\begin{figure}[t]\label{fig:lambdas}
\small
$\fbox{\inferrule{}{\sctx \vdash S : \psi}}$
~~~~$\sctx ::= \emptyset \pipe \sctx, x : \psi$
\begin{mathpar}
\inferrule[S-T-Stringin-I]
{s \in \lang{r}}
{\sctx \vdash \strin{s} : \stringin{r}}

@@

\inferrule[S-T-replace]
{\sctx \vdash S_1 : \stringin{r_1} \\ \sctx \vdash S_2 : \stringin{r_2} \\\\ {\tt lsubst}(r,r_1,r_2) = r'}
{\sctx \vdash \sisubst{r}{S_1}{S_2} : \stringin{r'}}

@@

\begin{figure}[t]
\small
$\fbox{\inferrule{}{S \sreduces S}}$
$\fbox{\inferrule{}{S \err}}$
\begin{mathpar}
\inferrule[S-E-RStr]
{ \ }
{\strin{s} \sreduces \strin{s}}

@@

\inferrule[S-E-Replace]
{S_1 \sreduces \strin{s_1} \\ S_2 \sreduces \strin{s_2} \\ \dosubst{r}{s_1}{s_2} = s} 
{\sisubst{r}{S_1}{S_2} \sreduces \sistr{s}}

@@

\begin{mathpar}
\inferrule[P-T-String]
{ \ }
{\tctx \vdash \tstr{s} : \str}

@@

\begin{mathpar}
\inferrule[P-E-Str]
{ \ }
{\tstr{s} \treduces \tstr{s}}

@@

\inferrule[P-E-Replace]
{P_1 \treduces \rx{r} \\ P_2 \treduces \tstr{s_2} \\ P_3 \treduces \tstr{s_3} \\ \dosubst{r}{s_2}{s_3} = s} 
{\preplace{P_1}{P_2}{P_3} \treduces \tstr{s}}

@@

Our type safety proof for language S replies on a relationship between
string substitution and language substitution given in lemma \ref{thm:substcorrespondence}.
We also rely upon several other properties of regular languages.
Throughout this section, we fix an alphabet $\Sigma$ over which strings $s$ and
regular expressions $r$ are defined. throughout the paper, $\lang{r}$ refers to the
language recognized by the expression $r$. This distinction between the expression
and its language -- typically elided in the literature -- makes our definition and
proofs about systems S and P more readable.

\begin{lem}{Properties of Regular Languages and Expressions.} \label{thm:regexprops}
The following are well-known properties of regular expressions which are necessary for our proofs:
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $s_1s_2 \in \lang{r_1r_2}$.
For all strings $s$ and expressions $r$, either $s \in \lang{r}$ or $s \not \in \lang{r}$.
Regular languages are closed under complements and concatenation.
The regular expressions correspond bijectively to the regular languages.
\end{lem}

\begin{defn}[$\tt{lsubst}$]
The function $\lsubst{r}{s_1}{s_2}$ produces a string in which all substrings of $s_1$ matching $r$ are replaced with $s_2$.
\end{defn}

\begin{defn}[$\tt{lreplace}$]
The function $\lreplace{r}{r_1}{r_2}$ produces a regular expression in which any sublanguage $\lang{r_1'}$ of $\lang{r_1}$ satisfying the condition $\lang{r_1'} \subseteq{\lang{r}}$ is replaced with $\lang{r_2}$.
\end{defn}

\begin{lem}{Closure and Totality of Replacement.} \label{thm:total}
If $r, r_1$ and $r_2$ are regular expressions, then $\lreplace{r}{r_1}{r_2}$ is also a regular expression.
\end{lem}
\begin{proof}
By induction on $r$ and closure properties of regular expressions.
\end{proof}

\begin{lem}{Substitution Correspondence.} \label{thm:substcorrespondence}
If $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ then $\lsubst{r}{s_1}{s_2} \in \lang{\lreplace{r}{r_1}{r_2}}$.
\end{lem}
\begin{proof}
The proof proceeds by structural induction on $r$.
The base case ($r=\alpha, \alpha \in \Sigma$) is trivial.
The choice and sequential cases require lemma 1, and prove a more general property that $\lsubst{r}{s_1}{s_2} \in \lang{r'}$ for some $r'$ isomorphic to $\lreplace{r}{r_1}{r_2}$.
The closure case follows from the induction hypothesis applied to any $n$-unwinding.

%\begin{description}
%\item[case $r=\alpha$.]
%If $s_1 = \alpha$ then $\alpha \in \lang{r_1}$ by assumption.
%Therefore, $\lsubst{r}{s_1}{s_2} = \lsubst{\alpha}{\alpha}{s_2} = s_2$ and
%$\lreplace{r}{r_1}{r_2} = \lreplace{\alpha}{r_1}{r_2}$. 
%Since $s_1 = \alpha$ and $\alpha \in \lang{r_1}$, $r_1 \iso \alpha | r_1'$ for some $r_1'$.
%Therefore, $\lreplace{\alpha}{r_1}{r_2} \iso \lreplace{\alpha}{\alpha|r_1'}{r_2}$ by Lemma X(1).
%Finally, $s_2 \in \lang{r_2}$ which implies $s_1 \in \lang{r_2|r'}$.
%If $s_1 \not = \alpha$ the $\lsubst{r}{s_1}{s_2} = s_1$ and $\lreplace{\alpha}{r_1}{r_2} = r_1$.
%
%\item[case $r = a|b$].
%Note that $\subst{a|b}{s_1}{s_2} = \lsubst{a}{\lsubst{b}{s_1}{s_2}}{s_2}$
%and $\lsubst{b}{s_1}{s_2} \in \lreplace{b}{r_1}{r_2}$ by induction.
%Therefore, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a}{\lreplace{b}{r_1}{r_2}}{r_2}$ by induction and the definition of lreplace.
%Finally, applying definitions once more, $\lsubst{a|b}{s_1}{s_2} \in \lreplace{a|b}{r_1}{r_2}$.
%
%\item[case $r = ab$].
%By a similar argument to the disjunctive case.
%
%\item[case $r = a*$].
%By considering the once unwinding of $a*$, noting that $s_1$ and $s_2$ are finite.
%\end{description}
\end{proof}

@@

\begin{lem}
  If $\sctx \vdash S : \stringin{r}$ then $r$ is a well-formed regular expression. 
\end{lem}
\begin{proof}
  The only non-trivial case is S-T-Replace, which follows from lemma \ref{thm:total}. 
\end{proof}

\begin{lem}
  If $\tctx \vdash P : \regex$ then $P \treduces \rx{r}$ such that $r$ is a well-formed regular expression. 
\end{lem}

@@

\begin{thm}{Safety for the String Fragment of P.}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \stringin{r}$ 
  then $S \sreduces \rstr{s}$ and $\rstr{s} : \stringin{r}$, or else $S \err$.
\end{thm}
\begin{proof}
By induction on the derivation of $\sctx \vdash S : \psi$. The interesting case
is S-T-Replace, which requires lemma \ref{thm:substcorrespondence}. The 
Coercion lemma additionally requires the second property of lemma \ref{thm:regexprops}.
